Appendix

### 2014-11-25 Rmarkdown/appendix.Rmd
Updated appendix

## CpGs are underrepresented in the human genome

This short R script demonstrates that CpGs are underrepresented in the human reference genome (build hg19).

@@

```{r dinucleotide_distribution, fig.lp = "fig:dinuclotide_distribution"}
library(ggplot2)
thesis_theme <- theme_classic(base_size = 12)
thesis_theme <- theme_bw(base_size = 12)
edf <- nf %*% t(nf) / (sum(nf %*% t(nf))) * 100
x <- data.frame(Dinucleotide = names(rowSums(df)),
                Observed = as.vector(100 * rowSums(df) / sum(as.numeric(df))), 
                Expected = as.vector(edf))
ggplot(x, aes(x = Dinucleotide, y = Observed / Expected)) + 
  geom_point(size = 5) + ylim(c(0, 1.5)) + 
  ggtitle("CpGs are underrepresented in human reference genome (hg19)") + 
  thesis_theme +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  geom_hline(yintercept = 1, linetype = 2)
```

>>>>

## CpGs are underrepresented in the human genome (hg19)

@@

```{r dinucleotide_distribution, fig.lp = "fig:dinuclotide_distribution"}
library(ggplot2)
thesis_theme <- theme_classic(base_size = 12)
thesis_theme <- theme_bw(base_size = 12)
edf <- nf %*% t(nf) / (sum(nf %*% t(nf))) * 100
x <- data.frame(Dinucleotide = names(rowSums(df)),
                Observed = as.vector(100 * rowSums(df) / sum(as.numeric(df))), 
                Expected = as.vector(edf))
ggplot(x, aes(x = Dinucleotide, y = Observed / Expected)) + 
  geom_point(size = 5) + coord_cartestian(ylim = c(0, 1.5)) + 
  ggtitle("CpGs are underrepresented in human reference genome (hg19)") + 
  thesis_theme +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  geom_hline(yintercept = 1, linetype = 2)
```

## Distances between CpGs in the human reference genome (hg19)

I extract the co-ordinates of all CpGs in the human reference genome (hg19) and compute the intra-pair distances (IPD) between adjacent CpGs. I then plot the empirical cumulative distribution function of IPD. These are compared to the distances we would expect if CpGs were uniformly distributed across the genome.

```{r cpg_ipds, cache = TRUE}
library(BSgenome.Hsapiens.UCSC.hg19)
set.seed(666)

# Observed IPDs
cpgs <- bsapply(new("BSParams", X = Hsapiens, FUN = matchPattern, 
                    exclude = c("rand", "Un", "hap"), simplify = TRUE),
                pattern = "CG")
d <- unlist(lapply(cpgs, start), use.names = FALSE)
ipd <- diff(d)
# Drop negative distances, which correspond to CpGs on different chromosomes
ipd <- ipd[ipd > 0]

# Expected IPDs if CpGs were uniformly distributed on genome
d_sim <- bsapply(new("BSParams", X = Hsapiens, FUN = function(x) {
  cpg_freq <- countPattern("CG", x)
  sort(sample(x = length(x), size = cpg_freq, replace = FALSE))
}, exclude = c("rand", "Un", "hap"), simplify = TRUE))
ipd_sim <- diff(unlist(d_sim, use.names = FALSE))
# Drop negative distances, which correspond to CpGs on different chromosomes
ipd_sim <- ipd_sim[ipd_sim > 0]

# Drop negative values, which are due to CpGs being on different chromosomes
ipd_df <- data.frame(IPD = c(ipd, ipd_sim), 
                  Source = c(rep("Observed", length(ipd)), 
                             rep("Expected", length(ipd_sim))))
```

### Empirical cumulative distribution plot

```{r cpg_ipds_cdf}
library(ggplot2)
thesis_theme <- theme_classic(base_size = 12)
thesis_theme <- theme_bw(base_size = 12)

ggplot(aes(x = IPD, colour = Source), data = ipd_df) + stat_ecdf() + thesis_theme + 
  coord_cartesian(xlim = c(0, 1000)) + 
  xlab("Distance between CpGs (bp)\n(Axis truncated at 1000)") + 
  ggtitle("Human reference genome (hg19)")
```

### Histogram

```{r cpg_ipds_hist}
library(ggplot2)
library(scales)
thesis_theme <- theme_classic(base_size = 12)
thesis_theme <- theme_bw(base_size = 12)

ipd_table <- table(ipd)
ipd_sim_table <- table(ipd_sim)
ipd_table_df <- data.frame(IPD = c(as.numeric(names(ipd_table)), 
                                   as.numeric(names(ipd_sim_table))), 
                           Count = c(ipd_table, ipd_sim_table), 
                           Source = c(rep("Observed", length(ipd_table)), 
                                      rep("Expected", length(ipd_sim_table))))

ggplot(aes(x = IPD, y = Count), 
       data = subset(ipd_table_df, Source == "Observed")) + 
  geom_line() + coord_cartesian(xlim = c(0, 300)) + 
  xlab("Distance between CpGs (bp)\n (Axis truncated at 300)") + 
  ggtitle("Human reference genome (hg19)") + 
  scale_y_continuous(labels = comma)
```
